---
title: "Dept. Of Defense Grant Application"
author: |
  | Stephen Pederson
  | Dame Roma Mitchell Cancer Research Laboratories
  | Adelaide Medical School
  | University of Adelaide
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

```{r packages}
pkg <- c(
  "tidyverse", "pzfx", "pander", "glue", "cowplot", "ggrepel", "parallel",
  "lme4", #"lmerTest",
  "reactable", "htmltools"
)
pacman::p_load(char = pkg, character.only = TRUE, update = FALSE)
```

This workflow requires the packages `r pander(c("workflowr", "pacman", "here", "lmerTest", pkg))`

```{r options}
theme_set(
  theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))
)
panderOptions("big.mark", ",")
panderOptions("table.split.table", Inf)
```


## Introduction

```{r suumarise_sheets}
div(
  class = "sheets",
  div(
    class = "sheets-hdr",
    em(
      class = "sheets-title",
      "All named sheets within the provided Graphpad Prism files. 
Any files specific to a single day or marked as a copy were excluded."
    )
  ),
  here::here("data/Tumour_power_calculation_data/") %>%
  list.files(full.names = TRUE) %>%
  lapply(
    function(x){
      tibble(file = x, `possible sheets` = pzfx_tables(x))
    }
  ) %>%
  bind_rows() %>%
  mutate(file = basename(file)) %>%
  dplyr::filter(
    !str_detect(`possible sheets`, "[Dd]ay|[Cc]opy"),
  ) %>%
  rename_all(str_to_title) %>%
  reactable::reactable(
    groupBy = "File", 
    columns = list(
      File = colDef(aggregate = "unique"),
      `Possible Sheets` = colDef(aggregate = "count")
    ),
    class = "sheets-tbl"
  )
)
```


Data was provided as above with most files being derived from [Hickey et al, 2021](https://www.nature.com/articles/s41591-020-01168-7).
File names reflect the figures and data files from this paper.

The models included were E418, HCI005, Gar15-13, UCD4, UCD65 and MH18 (not from the specified paper).
Only Vehicle, SARM and P4 treated samples were of interest.

## Assessment of Growth Rates {.tabset}

Each individual model will have a variable growth rate and these were first explored in order to gain an understanding of any associated variability.

### E418

```{r load_e418}
e418 <- here::here(
  "data/Tumour_power_calculation_data/", 
  "3b - E418 UCD4 E SARM DHT final data.pzfx"
  ) %>%
  read_pzfx(table = "Final growth data") %>%
  dplyr::rename(day = Var.1) %>%
  pivot_longer(
    cols = -day,
    names_to = "tumour",
    values_to = "volume"
  ) %>%
  dplyr::filter(!is.na(volume), volume > 0) %>%
  separate(
    tumour,
    into = c("treatment", "replicate"),
    sep = "_", 
    remove = FALSE
  ) %>%
  mutate(
    treatment = as.factor(treatment) %>% relevel(ref = "E2+PBS"),
    replicate = fct_inseq(replicate)
  ) %>%
  group_by(tumour) %>%
  mutate(n_points = dplyr::n()) %>%
  ungroup()
```

```{r plot_e418, fig.height=8, fig.width=10, fig.cap = "*All growth curves for individual tumours using the E418 model*"}
e418 %>%
  dplyr::filter(
    n_points == 9 # Only the complete sets of values
  ) %>%
  droplevels() %>%
  ggplot(
    aes(day, volume, colour = treatment)
  ) +
  geom_point() +
  geom_smooth(
    aes(group = replicate),
    method = "lm",
    se = FALSE
  ) +
  facet_wrap(~treatment + replicate, ncol = 10) +
  scale_y_log10() +
  labs(
    x = "Day",
    y = expression(paste("Tumour Volume (", mm^3, ")")),
    colour = "Treatment"
  )
```


```{r plot_resid_e418, fig.cap = "Residuals for E418 testing for A) linearity and B) homoscedasticity. The clear non-linearity for both treatment arms reflect the plateau in growth after about day 40."}
e418_lmer <- lmer(
  log2(volume) ~ treatment*day + (1|tumour),
  data = e418,
  subset = n_points == 9
) 
a <- e418_lmer@frame %>%
  mutate(
    resid = resid(e418_lmer, type = "pearson"),
    fitted = fitted(e418_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, resid)
  ) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~treatment, ncol = 1)
b <- e418_lmer@frame %>%
  mutate(
    resid = resid(e418_lmer, type = "pearson"),
    fitted = fitted(e418_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, abs(resid))
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment, ncol = 1) 
plot_grid(a, b, nrow = 1, labels = c("A", "B"))
```

```{r summary_e418}
summary(e418_lmer)$coef %>%
  as_tibble(rownames = "Coefficient") %>%
  pander(
    justify = "lrrr",
    caption = glue(
      "
      Mixed-effects output from the E418 PDX model.
      Initial tumour volumes were consistently higher for both treatment.
      Control PDX samples had a doubling time of {round(1/fixef(e418_lmer)['day'], 1)} days.
      This was approximately 1/3 slower in both treatment arms."
    )
  )
```

### HCI005

```{r load_hci005}
hci005 <- here::here(
  "data/Tumour_power_calculation_data", 
  "3b - HCI_005 15-13D Absolute tumor volumes.pzfx"
) %>%
  read_pzfx(table = "HCI005 Combined DHTEno AVG plots") %>%
  dplyr::rename(day = Var.1) %>%
  pivot_longer(
    cols = -day,
    names_to = "tumour",
    values_to = "volume"
  ) %>%
  dplyr::filter(!is.na(volume), volume > 0) %>%
  separate(
    tumour,
    into = c("treatment", "replicate"),
    sep = "_", 
    remove = FALSE
  ) %>%
  mutate(
    treatment = as.factor(treatment) %>% relevel(ref = "Veh"),
    replicate = fct_inseq(replicate)
  ) %>%
  group_by(tumour) %>%
  mutate(n_points = dplyr::n()) %>%
  ungroup() %>%
  mutate(
    volume = case_when(
      replicate == 3 & treatment == "DHT" & day == 28 ~ 100, # This was a clear typo
      TRUE ~ volume
    )
  )
```

```{r plot_hci005, fig.height=6, fig.width=8, fig.cap = "*All growth curves for individual tumours using the HCI005 model*"}
hci005 %>%
  dplyr::filter(
    n_points >= 8 # Only the complete sets of values
  ) %>%
  droplevels() %>%
  ggplot(
    aes(day, volume, colour = treatment)
  ) +
  geom_point() +
  geom_smooth(
    aes(group = replicate),
    method = "lm",
    se = FALSE
  ) +
  facet_wrap(~treatment + replicate, ncol = 7) +
  scale_y_log10() +
  labs(
    x = "Day",
    y = expression(paste("Tumour Volume (", mm^3, ")")),
    colour = "Treatment"
  )
```


```{r plot_resid_hci005, fig.cap = "Residuals for HCI005 testing for A) linearity and B) homoscedasticity. The clear non-linearity in the Vehicle arm is likely due to Replicate 5."}
hci005_lmer <- lmer(
  log2(volume) ~ treatment*day + (1|tumour),
  data = hci005,
  subset = n_points >= 8
) 
a <- hci005_lmer@frame %>%
  mutate(
    resid = resid(hci005_lmer, type = "pearson"),
    fitted = fitted(hci005_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, resid)
  ) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~treatment, ncol = 1, scales = "free_y")
b <- hci005_lmer@frame %>%
  mutate(
    resid = resid(hci005_lmer, type = "pearson"),
    fitted = fitted(hci005_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, abs(resid))
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment, ncol = 1) 
plot_grid(a, b, nrow = 1, labels = c("A", "B"))
```

```{r summary_hci005}
summary(hci005_lmer)$coef %>%
  as_tibble(rownames = "Coefficient") %>%
  pander(
    justify = "lrrr",
    caption = glue(
      "
      Mixed-effects output from the HCI005 PDX model.
      Initial tumour volumes were consistently higher for both treatment.
      Control PDX samples had a doubling time of {round(1/fixef(hci005_lmer)['day'], 1)} days.
      This was approximately zero in both treatment arms."
    )
  )
```


### Gar15-13 (NatMed 3f)

```{r load_gar15}
gar15 <- here::here(
  "data/Tumour_power_calculation_data", 
  "3f - Compilation of Gar15-13_PDX experiments.pzfx"
) %>%
  read_pzfx(table = "3F Combined Tumour Growth") %>%
  rename(day = Days) %>%
  pivot_longer(
    cols = -day,
    names_to = "tumour",
    values_to = "volume"
  ) %>%
  dplyr::filter(!is.na(volume), volume > 0) %>%
  separate(
    tumour,
    into = c("treatment", "replicate"),
    sep = "_", 
    remove = FALSE
  ) %>%
  mutate(
    treatment = as.factor(treatment) %>% relevel(ref = "Vehicle"),
    replicate = fct_inseq(replicate)
  ) %>%
  group_by(tumour) %>%
  mutate(n_points = dplyr::n()) %>%
  ungroup() 
```

```{r plot_gar15, fig.height=8, fig.width=10, fig.cap = "*All growth curves for individual tumours using the Gar15-13 model*"}
gar15 %>%
  dplyr::filter(
    n_points >= 8 # Only the complete sets of values
  ) %>%
  droplevels() %>%
  ggplot(
    aes(day, volume, colour = treatment)
  ) +
  geom_point() +
  geom_smooth(
    aes(group = replicate),
    method = "lm",
    se = FALSE
  ) +
  facet_wrap(~treatment + replicate, ncol = 8) +
  scale_y_log10() +
  labs(
    x = "Day",
    y = expression(paste("Tumour Volume (", mm^3, ")")),
    colour = "Treatment"
  )
```


```{r plot_resid_gar15, fig.cap = "Residuals for Gar15-13 testing for A) linearity and B) homoscedasticity."}
gar15_lmer <- lmer(
  log2(volume) ~ treatment*day + (1|tumour),
  data = gar15,
  subset = n_points >= 8
) 
a <- gar15_lmer@frame %>%
  mutate(
    resid = resid(gar15_lmer, type = "pearson"),
    fitted = fitted(gar15_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, resid)
  ) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~treatment, ncol = 1, scales = "free_y")
b <- gar15_lmer@frame %>%
  mutate(
    resid = resid(gar15_lmer, type = "pearson"),
    fitted = fitted(gar15_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, abs(resid))
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment, ncol = 1) 
plot_grid(a, b, nrow = 1, labels = c("A", "B"))
```

```{r summary_gar15}
summary(gar15_lmer)$coef %>%
  as_tibble(rownames = "Coefficient") %>%
  pander(
    justify = "lrrr",
    caption = glue(
      "
      Mixed-effects output from the Gar15-13 PDX model.
      Initial tumour volumes were consistently higher for both treatment.
      Control PDX samples had a doubling time of {round(1/fixef(gar15_lmer)['Days'], 1)} days.
      This was approximately doubled in DHT arm and extended by 30% in the Enobosarm arm."
    )
  )
```

### Gar15-13 (NatMed 4c)

```{r load_gar15_4c}
gar15_4c <- here::here(
  "data/Tumour_power_calculation_data", 
  "4c - CEL19-05_Gar1513 20191122.pzfx"
) %>%
  read_pzfx(table = "KeeMing_edited_1816_Growthcurves") %>%
  dplyr::rename(day = Var.1) %>%
  pivot_longer(
    cols = -day,
    names_to = "tumour",
    values_to = "volume"
  ) %>%
  dplyr::filter(!is.na(volume), volume > 0) %>%
  separate(
    tumour,
    into = c("treatment", "replicate"),
    sep = "_", 
    remove = FALSE
  ) %>%
  mutate(
    treatment = as.factor(treatment) %>% relevel(ref = "Con"),
    replicate = fct_inseq(replicate)
  ) %>%
  group_by(tumour) %>%
  mutate(n_points = dplyr::n()) %>%
  ungroup() 
```

```{r plot_gar15_4c, fig.height=8, fig.width=10, fig.cap = "*All growth curves for individual tumours using the Gar15-13 model*"}
gar15_4c %>%
  dplyr::filter(
    n_points >= 5 # Only the complete sets of values
  ) %>%
  droplevels() %>%
  ggplot(
    aes(day, volume, colour = treatment)
  ) +
  geom_point() +
  geom_smooth(
    aes(group = replicate),
    method = "lm",
    se = FALSE
  ) +
  facet_wrap(~treatment + replicate, ncol = 8) +
  scale_y_log10() +
  labs(
    x = "Day",
    y = expression(paste("Tumour Volume (", mm^3, ")")),
    colour = "Treatment"
  )
```


```{r plot_resid_gar15_4c, fig.cap = "Residuals for Gar15-13 testing for A) linearity and B) homoscedasticity."}
gar15_lmer_4c <- lmer(
  log2(volume) ~ treatment*day + (1|tumour),
  data = gar15_4c,
  subset = n_points >= 5
) 
a <- gar15_lmer_4c@frame %>%
  mutate(
    resid = resid(gar15_lmer_4c, type = "pearson"),
    fitted = fitted(gar15_lmer_4c)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, resid)
  ) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~treatment, ncol = 1, scales = "free_y")
b <- gar15_lmer_4c@frame %>%
  mutate(
    resid = resid(gar15_lmer_4c, type = "pearson"),
    fitted = fitted(gar15_lmer_4c)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, abs(resid))
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment, ncol = 1) 
plot_grid(a, b, nrow = 1, labels = c("A", "B"))
```

```{r summary_gar15_4c}
summary(gar15_lmer_4c)$coef %>%
  as_tibble(rownames = "Coefficient") %>%
  pander(
    justify = "lrrr",
    caption = glue(
      "
      Mixed-effects output from the Gar15-13 PDX model.
      Control PDX samples had a doubling time of {round(1/fixef(gar15_lmer_4c)['day'], 1)} days.
      This was increased significantly in both Enobosarm arms."
    )
  )
```

### UCD4

```{r load_ucd4}
ucd4 <- here::here(
  "data/Tumour_power_calculation_data/", 
  "ED9b - UCD4 growth and mass data compiled.pzfx"
  ) %>%
  read_pzfx(table = "E66,70 combined growth") %>%
  dplyr::rename(day = days) %>%
  pivot_longer(
    cols = -day,
    names_to = "tumour",
    values_to = "volume"
  ) %>%
  dplyr::filter(
    !is.na(volume), volume > 0,
    !str_detect(tumour, "placebo")
  ) %>%
  separate(
    tumour,
    into = c("treatment", "replicate"),
    sep = "_", 
    remove = FALSE
  ) %>%
  mutate(
    treatment = as.factor(treatment) %>% relevel(ref = "E"),
    replicate = fct_inseq(replicate)
  ) %>%
  group_by(tumour) %>%
  mutate(n_points = dplyr::n()) %>%
  ungroup()
```

```{r plot_ucd4, fig.height=8, fig.width=10, fig.cap = "*All growth curves for individual tumours using the UCD4 model*"}
ucd4 %>%
  dplyr::filter(
    n_points >= 6 # Only the complete sets of values
  ) %>%
  droplevels() %>%
  ggplot(
    aes(day, volume, colour = treatment)
  ) +
  geom_point() +
  geom_smooth(
    aes(group = replicate),
    method = "lm",
    se = FALSE
  ) +
  facet_wrap(~treatment + replicate, ncol = 9) +
  scale_y_log10() +
  labs(
    x = "Day",
    y = expression(paste("Tumour Volume (", mm^3, ")")),
    colour = "Treatment"
  )
```


```{r plot_resid_ucd4, fig.cap = "Residuals for UCD4 testing for A) linearity and B) homoscedasticity."}
ucd4_lmer <- lmer(
  log2(volume) ~ treatment*day + (1|tumour),
  data = ucd4,
  subset = n_points >= 6
) 
a <- ucd4_lmer@frame %>%
  mutate(
    resid = resid(ucd4_lmer, type = "pearson"),
    fitted = fitted(ucd4_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, resid)
  ) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~treatment, ncol = 1)
b <- ucd4_lmer@frame %>%
  mutate(
    resid = resid(ucd4_lmer, type = "pearson"),
    fitted = fitted(ucd4_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, abs(resid))
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment, ncol = 1) 
plot_grid(a, b, nrow = 1, labels = c("A", "B"))
```

```{r summary_ucd4}
summary(ucd4_lmer)$coef %>%
  as_tibble(rownames = "Coefficient") %>%
  pander(
    justify = "lrrr",
    caption = glue(
      "
      Mixed-effects output from the UCD4 PDX model.
      Initial tumour volumes were consistently higher for both treatment.
      Control PDX samples had a doubling time of {round(1/fixef(ucd4_lmer)['day'], 1)} days.
      This was approximately 1/3 slower in both treatment arms."
    )
  )
```

### UCD65 (ED9b)

```{r load_ucd65}
ucd65 <- here::here(
  "data/Tumour_power_calculation_data", 
  "ED9b - UCD65 growth and mass data compiled.pzfx"
) %>%
  read_pzfx(
    table = "Tumor volume E E+P plac P"
  ) %>%
  dplyr::rename(day = Var.1) %>%
  pivot_longer(
    cols = -day,
    names_to = "tumour",
    values_to = "volume"
  ) %>%
  dplyr::filter(
    !is.na(volume), volume > 0,
    !str_detect(tumour, "placebo")
  ) %>%
  separate(
    tumour,
    into = c("treatment", "replicate"),
    sep = "_", 
    remove = FALSE
  ) %>%
  mutate(
    treatment = as.factor(treatment) %>% relevel(ref = "E"),
    replicate = fct_inseq(replicate)
  ) %>%
  group_by(tumour) %>%
  mutate(n_points = dplyr::n()) %>%
  ungroup()
```

```{r plot_ucd65, fig.height=6, fig.width=8, fig.cap = "*All growth curves for individual tumours using the UCD65 model*"}
ucd65 %>%
  dplyr::filter(
    n_points >= 6 # Only the complete sets of values
  ) %>%
  droplevels() %>%
  ggplot(
    aes(day, volume, colour = treatment)
  ) +
  geom_point() +
  geom_smooth(
    aes(group = replicate),
    method = "lm",
    se = FALSE
  ) +
  facet_wrap(~treatment + replicate, ncol = 5) +
  scale_y_log10() +
  labs(
    x = "Day",
    y = expression(paste("Tumour Volume (", mm^3, ")")),
    colour = "Treatment"
  )
```


```{r plot_resid_ucd65, fig.cap = "Residuals for UCD65 testing for A) linearity and B) homoscedasticity."}
ucd65_lmer <- lmer(
  log2(volume) ~ treatment*day + (1|tumour),
  data = ucd65,
  subset = n_points >= 6
) 
a <- ucd65_lmer@frame %>%
  mutate(
    resid = resid(ucd65_lmer, type = "pearson"),
    fitted = fitted(ucd65_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, resid)
  ) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~treatment, ncol = 1)
b <- ucd65_lmer@frame %>%
  mutate(
    resid = resid(ucd65_lmer, type = "pearson"),
    fitted = fitted(ucd65_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, abs(resid))
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment, ncol = 1) 
plot_grid(a, b, nrow = 1, labels = c("A", "B"))
```

```{r summary_ucd65}
summary(ucd65_lmer)$coef %>%
  as_tibble(rownames = "Coefficient") %>%
  pander(
    justify = "lrrr",
    caption = glue(
      "
      Mixed-effects output from the UCD65 PDX model.
      Initial tumour volumes were consistently higher for both treatment.
      Control PDX samples had a doubling time of {round(1/fixef(ucd65_lmer)['day'], 1)} days.
      This was one day slower in both treatment arms."
    )
  )
```

### UCD65 (ED9c)

```{r load_ucd65_9c}
ucd65_9c <- here::here(
  "data/Tumour_power_calculation_data", 
  "ED9c - E434 UCD65 E SARM P4 final data.pzfx"
) %>%
  read_pzfx(
    table = "Final groups volume"
  ) %>%
  dplyr::rename(day = Var.1) %>%
  pivot_longer(
    cols = -day,
    names_to = "tumour",
    values_to = "volume"
  ) %>%
  dplyr::filter(
    !is.na(volume), volume > 0
  ) %>%
  separate(
    tumour,
    into = c("treatment", "replicate"),
    sep = "_", 
    remove = FALSE
  ) %>%
  mutate(
    treatment = as.factor(treatment) %>% relevel(ref = "E2+PBS"),
    replicate = fct_inseq(replicate)
  ) %>%
  group_by(tumour) %>%
  mutate(n_points = dplyr::n()) %>%
  ungroup()
```

```{r plot_ucd65_9c, fig.height=6, fig.width=8, fig.cap = "*All growth curves for individual tumours using the UCD65 model*"}
ucd65_9c %>%
  dplyr::filter(
    n_points >= 8 # Only the complete sets of values
  ) %>%
  droplevels() %>%
  ggplot(
    aes(day, volume, colour = treatment)
  ) +
  geom_point() +
  geom_smooth(
    aes(group = replicate),
    method = "lm",
    se = FALSE
  ) +
  facet_wrap(~treatment + replicate, ncol = 5) +
  scale_y_log10() +
  labs(
    x = "Day",
    y = expression(paste("Tumour Volume (", mm^3, ")")),
    colour = "Treatment"
  )
```


```{r plot_resid_ucd65_9c, fig.cap = "Residuals for UCD65 testing for A) linearity and B) homoscedasticity."}
ucd65_lmer_9c <- lmer(
  log2(volume) ~ treatment*day + (1|tumour),
  data = ucd65_9c,
  subset = n_points >= 8
) 
a <- ucd65_lmer_9c@frame %>%
  mutate(
    resid = resid(ucd65_lmer_9c, type = "pearson"),
    fitted = fitted(ucd65_lmer_9c)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, resid)
  ) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~treatment, ncol = 1)
b <- ucd65_lmer_9c@frame %>%
  mutate(
    resid = resid(ucd65_lmer_9c, type = "pearson"),
    fitted = fitted(ucd65_lmer_9c)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, abs(resid))
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment, ncol = 1) 
plot_grid(a, b, nrow = 1, labels = c("A", "B"))
```

```{r summary_ucd65_9c}
summary(ucd65_lmer_9c)$coef %>%
  as_tibble(rownames = "Coefficient") %>%
  pander(
    justify = "lrrr",
    caption = glue(
      "
      Mixed-effects output from the UCD65 PDX model.
      Initial tumour volumes were consistently higher for both treatment.
      Control PDX samples had a doubling time of {round(1/fixef(ucd65_lmer)['day'], 1)} days.
      This was slower in both treatment arms."
    )
  )
```

### MH18

```{r load_mh18}
mh18 <- here::here(
  "data/Tumour_power_calculation_data", 
  "MH18 tumour growth curves.pzfx"
) %>%
  read_pzfx(
    table = "Raw tumour volumes"
  ) %>%
  dplyr::rename(day = `Days from treatment commencement`) %>%
  pivot_longer(
    cols = -day,
    names_to = "tumour",
    values_to = "volume"
  ) %>%
  dplyr::filter(
    !is.na(volume), volume > 0,
    !str_detect(tumour, "Enzalutamide")
  ) %>%
  separate(
    tumour,
    into = c("treatment", "replicate"),
    sep = "_", 
    remove = FALSE
  ) %>%
  mutate(
    treatment = as.factor(treatment) %>% relevel(ref = "Vehicle"),
    replicate = fct_inseq(replicate)
  ) %>%
  group_by(tumour) %>%
  mutate(n_points = dplyr::n()) %>%
  ungroup()
```

```{r plot_mh18, fig.height=6, fig.width=8, fig.cap = "*All growth curves for individual tumours using the MH18 model*"}
mh18 %>%
  dplyr::filter(
    n_points >= 8 # Only the complete sets of values
  ) %>%
  droplevels() %>%
  ggplot(
    aes(day, volume, colour = treatment)
  ) +
  geom_point() +
  geom_smooth(
    aes(group = replicate),
    method = "lm",
    se = FALSE
  ) +
  facet_wrap(~treatment + replicate, ncol = 5) +
  scale_y_log10() +
  labs(
    x = "Day",
    y = expression(paste("Tumour Volume (", mm^3, ")")),
    colour = "Treatment"
  )
```


```{r plot_resid_mh18, fig.cap = "Residuals for MH18 testing for A) linearity and B) homoscedasticity."}
mh18_lmer <- lmer(
  log2(volume) ~ treatment*day + (1|tumour),
  data = mh18,
  subset = n_points >= 8
) 
a <- mh18_lmer@frame %>%
  mutate(
    resid = resid(mh18_lmer, type = "pearson"),
    fitted = fitted(mh18_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, resid)
  ) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~treatment, ncol = 1)
b <- mh18_lmer@frame %>%
  mutate(
    resid = resid(mh18_lmer, type = "pearson"),
    fitted = fitted(mh18_lmer)
  ) %>%
  as_tibble() %>%
  ggplot(
    aes(fitted, abs(resid))
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment, ncol = 1) 
plot_grid(a, b, nrow = 1, labels = c("A", "B"))
```

```{r summary_mh18}
summary(mh18_lmer)$coef %>%
  as_tibble(rownames = "Coefficient") %>%
  pander(
    justify = "lrrr",
    caption = glue(
      "
      Mixed-effects output from the MH18 PDX model.
      Initial tumour volumes were consistently higher for both treatment.
      Control PDX samples had a doubling time of {round(1/fixef(mh18_lmer)['day'], 1)} days.
      A near infinite doubling time was observed in the DHT and Enobosarm treatment arms."
    )
  )
```

## Relationship Between Variables {.tabset}

### Control Samples

```{r allControls}
allControls <- list(
  tibble(
    data = list(
      e418 %>%
        dplyr::filter(
          treatment == "E2+PBS",
          n_points >= 9
        ) %>%
        droplevels()
    ),
    pdx = "E418",
    control = "E2+PBS"
  ),
  tibble(
    data = list(
      hci005 %>%
        dplyr::filter(
          treatment == "Veh",
          n_points >= 8
        ) %>%
        droplevels()
    ),
    pdx = "HCI005",
    control = "Vehicle"
  ),
  tibble(
    data = list(
      gar15 %>%
        dplyr::filter(
          treatment == "Vehicle",
          n_points >= 8
        ) %>%
        droplevels()
    ),
    pdx = "Gar15-13 (3f)",
    control = "Vehicle"
  ),
  tibble(
    data = list(
      gar15_4c %>%
        dplyr::filter(
          treatment == "Con",
          n_points >= 5
        ) %>%
        droplevels()
    ),
    pdx = "Gar15-13 (4c)",
    control = "Vehicle"
  ),
  tibble(
    data = list(
      ucd4 %>%
        dplyr::filter(
          treatment == "E",
          n_points >= 6
        ) %>%
        droplevels()
    ),
    pdx = "UCD4",
    control = "E2"
  ),
  tibble(
    data = list(
      ucd65 %>%
        dplyr::filter(
          treatment == "E",
          n_points >= 6
        ) %>%
        droplevels()
    ),
    pdx = "UCD65 (ED9b)",
    control = "E2"
  ),
  tibble(
    data = list(
      ucd65_9c %>%
        dplyr::filter(
          treatment == "E2+PBS",
          n_points >= 8
        ) %>%
        droplevels()
    ),
    pdx = "UCD65 (ED9c)",
    control = "E2+PBS"
  ),
  tibble(
    data = list(
      mh18 %>%
        dplyr::filter(
          treatment == "Vehicle",
          n_points >= 8
        ) %>%
        droplevels()
    ),
    pdx = "MH18",
    control = "Vehicle"
  )
) %>%
  bind_rows() %>%
  mutate(
    coef = lapply(data, function(x){
      x %>%
        split(f = .$replicate) %>% 
        lapply(function(y){lm(log2(volume) ~ day, data = y)}) %>% 
        lapply(broom::tidy) %>% 
        bind_rows() %>% 
        group_by(term) %>% 
        summarise(
          est = mean(estimate), 
          sd = sd(estimate),
          .groups = "drop"
        )
    }),
    sigma = vapply(data, function(x){
      lm(log2(volume) ~ day*replicate, data = x) %>%
        resid() %>%
        sd
    },
    numeric(1))
  )
```

```{r tab_allControls}
allControls %>%
  unnest(coef) %>%
  pivot_wider(id_cols = c(data, pdx, control, term), names_from = term, values_from = c(est, sd)) %>%
  dplyr::select(
    PDX = pdx, 
    Control = control,
    `Mean Volume~0~` = `est_(Intercept)`,
    `SD Volume~0~` = `sd_(Intercept)`,
    `Mean Daily Growth` = est_day,
    `SD Daily Growth` = sd_day
  ) %>%
  arrange(desc(`Mean Daily Growth`)) %>%
  pander(
    justify = "llrrrr",
    caption = "
    Summary of values for initial tumour volume and how much this varies for each PDX replicate,
    along with the daily growth rate (i.e. slope) and how much this varies between replicates.
    Values are taken from the controls from each dataset, with no treated values.
    All values are shown in the log~2~ scale.
    "
  )
```

```{r plot_allControls, fig.width=10, fig.height=8, fig.cap = "*Relation ship between inter-replicate variability and both the daily growth rate and the initial volume. Slower growing tumours (Vehicle controls) were more consistent in initial size and were initially larger, whilst most faster growing tumours (E2 controls) varied noticeably in initial size. No relationship was found between the varibility in growth rate and the actual growth rate*"}
allControls %>%
  unnest(coef) %>%
  mutate(
    term = str_replace(term, "\\(Intercept\\)", "Initial Volume"),
    term = str_replace(term, "day", "Daily Growth"),
    pdx = str_replace(pdx, "\\(", "\n(")
  ) %>%
  ggplot(aes(est, sd)) +
  geom_text(
    aes(colour = control, label = pdx)
  ) +
  geom_smooth(method = MASS::rlm) +
  facet_wrap(~term, scales = "free") +
  scale_x_continuous(expand = expansion(0.08)) +
  labs(
    x = "Estimate",
    y = "Replicate SD",
    colour = "Controls"
  )
  
```

Given the above, data should be simulated for E2 and Vehicle controls separately

### SARM

```{r allSarm}
allSarm <- list(
  E418 = e418 %>% 
    mutate(pdx = "E418", control = "E2+PBS"), 
  HCI005 = hci005 %>% 
    mutate(pdx = "HCI005", control = "Vehicle"), 
  Gar15_3f = gar15 %>% 
    mutate(pdx = "Gar15-13 (3f)", control = "Vehicle"), 
  Gar15_4c = gar15_4c %>% 
    mutate(pdx = "Gar15-13 (4c)", control = "Vehicle"), 
  # UCD4 = ucd4, 
  # UCD65_9b = ucd65, 
  UCD65_9c = ucd65_9c %>%
    mutate(pdx = "UCD65 (ED9c)", control = "E2+PBS"), 
  MH18 = mh18 %>% 
    mutate(pdx = "MH18", control = "Vehicle")
) %>%
  lapply(
    dplyr::filter,
    str_detect(treatment, "(SARM|Eno)")
  ) %>%
  lapply(droplevels) %>%
  lapply(nest, data = -any_of(c("pdx", "control"))) %>%
  bind_rows() %>%
  mutate(
    coef = lapply(data, function(x){
      x %>%
        split(f = .$replicate) %>% 
        lapply(function(y){lm(log2(volume) ~ day, data = y)}) %>% 
        lapply(broom::tidy) %>%
        bind_rows() %>%
        group_by(term) %>%
        summarise(
          est = mean(estimate),
          sd = sd(estimate),
          .groups = "drop"
        )
    }),
    sigma = vapply(data, function(x){
      lm(log2(volume) ~ day*replicate, data = x) %>%
        resid() %>%
        sd
    },
    numeric(1))
  )
```

```{r tab_allSarm}
allSarm %>%
  unnest(coef) %>%
  pivot_wider(id_cols = c(data, pdx, control, term), names_from = term, values_from = c(est, sd)) %>%
  dplyr::select(
    PDX = pdx, 
    Control = control,
    `Mean Volume~0~` = `est_(Intercept)`,
    `SD Volume~0~` = `sd_(Intercept)`,
    `Mean Daily Growth` = est_day,
    `SD Daily Growth` = sd_day
  ) %>%
  arrange(desc(`Mean Daily Growth`)) %>%
  pander(
    justify = "llrrrr",
    caption = "
    Summary of values for initial tumour volume and how much this varies for each PDX replicate,
    along with the daily growth rate (i.e. slope) and how much this varies between replicates.
    Values are taken from the controls from each dataset, with no treated values.
    All values are shown in the log~2~ scale.
    "
  )
```

```{r plot_allSarm, fig.width=10, fig.height=8, fig.cap = "*Relation ship between inter-replicate variability and both the daily growth rate and the initial volume. Slower growing tumours (Vehicle controls) were more consistent in initial size and were initially larger, whilst most faster growing tumours (E2 controls) varied noticeably in initial size. No relationship was found between the varibility in growth rate and the actual growth rate*"}
allSarm %>%
  unnest(coef) %>%
  mutate(
    term = str_replace(term, "\\(Intercept\\)", "Initial Volume"),
    term = str_replace(term, "day", "Daily Growth"),
    pdx = str_replace(pdx, "\\(", "\n(")
  ) %>%
  ggplot(aes(est, sd)) +
  geom_text(
    aes(colour = control, label = pdx)
  ) +
  geom_smooth(method = MASS::rlm) +
  facet_wrap(~term, scales = "free") +
  scale_x_continuous(expand = expansion(0.08)) +
  labs(
    x = "Estimate",
    y = "Replicate SD",
    colour = "Controls"
  )
  
```

### Progesterone

```{r allPro}
allPro <- list(
  # E418 = e418 %>% 
  #   mutate(pdx = "E418", control = "E2+PBS"), 
  # HCI005 = hci005 %>% 
  #   mutate(pdx = "HCI005", control = "Vehicle"), 
  # Gar15_3f = gar15 %>% 
  #   mutate(pdx = "Gar15 (3f)", control = "Vehicle"), 
  # Gar15_4c = gar15_4c %>% 
  #   mutate(pdx = "Gar15 (4c)", control = "Vehicle"), 
  UCD4 = ucd4 %>%
    mutate(pdx = "UCD4", control = "E2"),
  UCD65_9b = ucd65 %>%
    mutate(pdx = "UCD65 (ED9b)", control = "E2"),
  UCD65_9c = ucd65_9c %>%
    mutate(pdx = "UCD65 (ED9c)", control = "E2+PBS")
  # MH18 = mh18 %>% 
  #   mutate(pdx = "MH18", control = "Vehicle")
) %>%
  lapply(
    dplyr::filter,
    str_detect(treatment, "P")
  ) %>%
  lapply(droplevels) %>%
  lapply(nest, data = -any_of(c("pdx", "control"))) %>%
  bind_rows() %>%
  mutate(
    coef = lapply(data, function(x){
      x %>%
        split(f = .$replicate) %>% 
        lapply(function(y){lm(log2(volume) ~ day, data = y)}) %>% 
        lapply(broom::tidy) %>%
        bind_rows() %>%
        group_by(term) %>%
        summarise(
          est = mean(estimate),
          sd = sd(estimate),
          .groups = "drop"
        )
    }),
    sigma = vapply(data, function(x){
      lm(log2(volume) ~ day*replicate, data = x) %>%
        resid() %>%
        sd
    },
    numeric(1))    
  )
```

```{r tab_allPro}
allPro %>%
  unnest(coef) %>%
  pivot_wider(id_cols = c(data, pdx, control, term), names_from = term, values_from = c(est, sd)) %>%
  dplyr::select(
    PDX = pdx, 
    Control = control,
    `Mean Volume~0~` = `est_(Intercept)`,
    `SD Volume~0~` = `sd_(Intercept)`,
    `Mean Daily Growth` = est_day,
    `SD Daily Growth` = sd_day
  ) %>%
  arrange(desc(`Mean Daily Growth`)) %>%
  pander(
    justify = "llrrrr",
    caption = "
    Summary of values for initial tumour volume and how much this varies for each PDX replicate,
    along with the daily growth rate (i.e. slope) and how much this varies between replicates.
    Values are taken from the controls from each dataset, with no treated values.
    All values are shown in the log~2~ scale.
    "
  )
```

```{r plot_allPro, fig.width=10, fig.height=8, fig.cap = "*Relation ship between inter-replicate variability and both the daily growth rate and the initial volume. Slower growing tumours (Vehicle controls) were more consistent in initial size and were initially larger, whilst most faster growing tumours (E2 controls) varied noticeably in initial size. No relationship was found between the varibility in growth rate and the actual growth rate*"}
allPro %>%
  unnest(coef) %>%
  mutate(
    term = str_replace(term, "\\(Intercept\\)", "Initial Volume"),
    term = str_replace(term, "day", "Daily Growth"),
    pdx = str_replace(pdx, "\\(", "\n(")
  ) %>%
  ggplot(aes(est, sd)) +
  geom_text(
    aes(colour = control, label = pdx)
  ) +
  geom_smooth(method = MASS::rlm) +
  facet_wrap(~term, scales = "free") +
  scale_x_continuous(expand = expansion(0.08)) +
  labs(
    x = "Estimate",
    y = "Replicate SD",
    colour = "Controls"
  )
```


### Combined

```{r plotCombined, fig.width=10, fig.height=8, fig.cap = "*All data combined across PDX lines, controls and treatments. NO relationship was found between variability of Growth rate and the estiamted growth rate, whilst for tumours with higher initial volumes, varibility betwene intial volumes was considerably lower.*"}
list(
  allControls %>% mutate(treatment = "Control"),
  allSarm %>% mutate(treatment = "Enobosarm"),
  allPro %>% mutate(treatment = "Progesterone")
) %>%
  bind_rows() %>%
  unnest(coef) %>%
  mutate(
    term = str_replace(term, "\\(Intercept\\)", "Initial Volume"),
    term = str_replace(term, "day", "Daily Growth"),
    pdx = str_replace(pdx, "\\(", "\n(")
  ) %>%
  ggplot(aes(est, sd)) +
  geom_text_repel(
    aes(colour = control, label = pdx),
    show.legend = FALSE
  ) +
  geom_point(
    aes(colour = control, shape = treatment),
    size = 3
  ) +
  geom_smooth(method = MASS::rlm) +
  facet_wrap(~term, scales = "free") +
  scale_x_continuous(expand = expansion(0.08)) +
  labs(
    x = "Estimate",
    y = "Replicate SD",
    colour = "Controls",
    shape = "Treatment"
  ) 
```

```{r plotSigma}
list(
  allControls %>% mutate(treatment = "Control"),
  allSarm %>% mutate(treatment = "Enobosarm"),
  allPro %>% mutate(treatment = "Progesterone")
) %>%
  bind_rows() %>% 
  mutate(pdx = str_replace(pdx, " \\(", "\n(")) %>%
  ggplot(aes(pdx, sigma, shape = treatment, colour = control)) + 
  geom_point(size = 3) +
  labs(
    x = "PDX",
    y = expression(sigma),
    shape = "Treatment",
    colour = "Controls"
  )
```

##

### Summary

As such, a *variability in the growth rate* was chosen to be 0.025 as this represented the 90th percentile of the non-UCD65 (ED9c) experiments, and if including this more variable experiment, this still is approximately equivalent to the 75th percentile.
This will also be a conservative over-estimate of the variability for the SARM-treated and highly responding lines.

For *initial tumour volume*, a value of 8 with inter-replicate variability of 0.4 will be appropriately conservative for those using a Vehicle control.
However, for those using an E2 control, an initial volume of 3 with variability of 1.2 will again provide a suitably conservative value.

For overall 'noise', a value of 0.25 will be representative for control samples, whilst this should be raised to 0.4 for treated tumours

## Power Curves

```{r}
library(lmerTest)
```

```{r simulateData}
simulateData <- function(n_control = 8, n_treat = 8, mn0 = 8, sd0 = 0.4, gr_control = 0.08, gr_sd = 0.025, gr_reduction = 0.3, sigma_control = 0.25, sigma_treat = 0.4, ...){
  
  ## Simulate the intercepts
  int_cont <- rnorm(n_control, mn0, sd0)
  int_treat <- rnorm(n_treat, mn0, sd0)
  
  ## Simulate the slopes
  slope_cont <- rnorm(n_control, gr_control, gr_sd)
  slope_treat <- rnorm(n_treat, (1-gr_reduction)*gr_control, gr_sd)
  
  ## Assume measurements every 7 days, for 8 weeks: 56 days
  days <- seq(0, 60, by = 7)
  
  tibble(
    replicate = c(seq_len(n_control), seq_len(n_treat)),
    intercept = c(int_cont, int_treat),
    slope = c(slope_cont, slope_treat),
    sd = c(rep(sigma_control, n_control), rep(sigma_treat, n_treat)),
    treatment = c(rep("control", n_control), rep("treated", n_treat)),
    sample = paste(treatment, replicate, sep = "_"),
    days = lapply(seq_len(n_control + n_treat), function(x){days})
  ) %>%
    unnest(days) %>%
    mutate(log2_volume = rnorm(nrow(.), intercept + slope*days, sd))
  
}
```

```{r alpha}
alpha <- 0.01
```


In all calculations below a significance threshold of &alpha; = `r alpha` was used.
**This allows for multiple testing for up to 5 experiments using Bonferroni's adjustment**.
Any increase in the number of experiments will require a more stringent threshold again.

For simplicity, the same number of replicates were assumed to be performed in both treated and control arms of the experiments.

### Vehicle Controls


```{r veh_res}
n_sim <- 1000
reductions <- c(0.2, 0.25, 0.3, 0.35, 0.4, 0.5)
veh_res <- list(
  tibble(
    nc = 6,
    nt = 6,
    reduction = reductions,
    n_rep = n_sim
  ),
    tibble(
    nc = 8,
    nt = 8,
    reduction = reductions,
    n_rep = n_sim
  ),
    tibble(
    nc = 10,
    nt = 10,
    reduction = reductions,
    n_rep = n_sim
  )
) %>%
  bind_rows() %>%
  split(f = seq_len(nrow(.))) %>%
  mclapply(
    mutate,
    n_sig = replicate(
      n_rep,
      {
        mod <- suppressWarnings(
          lmer(
            log2_volume ~ days*treatment + (1|sample), 
            data = simulateData(n_control = nc, n_treat = nt, gr_reduction = reduction)
          )
        )
        coef(summary(mod))["days:treatmenttreated", "Pr(>|t|)"] < alpha
      }
    ) %>%
      sum(),
    power = n_sig / n_rep,
    mc.cores = 6
  ) %>%
  bind_rows()
```

```{r plot_veh_res, fig.width=8, fig.height=6, fig.cap = "*Power curve for treatments using Vehicle controls. These tend to start with higher initial tumour volumes and lower varibility between replicates*"}
veh_res %>%
  mutate(
    se = sqrt(power*(1-power)/n_rep)
  ) %>%
  ggplot(
    aes(reduction, power, colour = as.factor(nc))
  ) +
  geom_point() +
  geom_errorbar(
    aes(ymin = power - se, ymax = power + se),
    width = 0.01
  ) +
  geom_line() +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Reduction in Growth Rate",
    y = "Power",
    colour = "Replicates"
  ) +
  ggtitle("Vehicle Controls")
```


### E2 Controls


```{r e2_res}
n_sim <- 1000
e2_res <- list(
  tibble(
    nc = 6,
    nt = 6,
    reduction = reductions,
    n_rep = n_sim
  ),
    tibble(
    nc = 8,
    nt = 8,
    reduction = reductions,
    n_rep = n_sim
  ),
    tibble(
    nc = 10,
    nt = 10,
    reduction = reductions,
    n_rep = n_sim
  )
) %>%
  bind_rows() %>%
  split(f = seq_len(nrow(.))) %>%
  mclapply(
    mutate,
    n_sig = replicate(
      n_rep,
      {
        mod <- suppressWarnings(
          lmer(
            log2_volume ~ days*treatment + (1|sample), 
            data = simulateData(n_control = nc, n_treat = nt, gr_reduction = reduction, mn0 = 3, sd0 = 1.2)
          )
        )
        coef(summary(mod))["days:treatmenttreated", "Pr(>|t|)"] < alpha
      }
    ) %>%
      sum(),
    power = n_sig / n_rep,
    mc.cores = 6
  ) %>%
  bind_rows()
```

```{r plot_e2_res, fig.width=8, fig.height=6, fig.cap = "*Power curve for treatments using E2 controls. These tend to start with lower initial tumour volumes and higher varibility between replicates*"}
e2_res %>%
  mutate(
    se = sqrt(power*(1-power)/n_rep)
  ) %>%
  ggplot(
    aes(reduction, power, colour = as.factor(nc))
  ) +
  geom_point() +
  geom_errorbar(
    aes(ymin = power - se, ymax = power + se),
    width = 0.01
  ) +
  geom_line() +
  scale_x_continuous(labels = scales::percent) +
  labs(
    x = "Reduction in Growth Rate",
    y = "Power",
    colour = "Replicates"
  ) +
  ggtitle(
    "E2 Controls"
  )
```




